---
title: "MRP Weighting for CID Consensus Screening"
author: "TreeSearch development"
date: today
format:
  html:
    toc: true
    code-fold: true
    self-contained: true
execute:
  cache: true
  warning: false
---

## Motivation

`InfoConsensus()` uses a dual-layer scoring architecture: a fast Fitch
parsimony layer on MRP (Matrix Representation with Parsimony) characters
screens TBR candidates, and a slower but exact CID (Clustering Information
Distance) layer accepts or rejects moves.  The tighter the parsimony
proxy tracks CID, the fewer wasted full-CID evaluations.

This analysis asks: **which parsimony weighting on the MRP characters best
correlates with CID among single-TBR neighbours?**

We compare:

- Equal weights (EW)
- Implied weights with $k \in \{1, 2, 3, 5, 10, 20\}$
- Profile parsimony

## Data

We use a 60-tip subsample of the mammals bootstrap dataset from
Lemoine et al. (2018) (1000 bootstrap trees on 1449 taxa), as bundled in the
`consense` package.  100 bootstrap trees form the reference pool.

```{r}
#| label: setup
#| message: false
library(TreeSearch)
library(TreeTools)
library(TreeDist)
library(phangorn)
library(ggplot2)
```

```{r}
#| label: load-data
mammal_trees <- ape::read.tree(
  "../../../consense/inst/extdata/phylocrisp/extracted/mammals_full/full/full_bootstraps.tre"
)

set.seed(4917)
all_tips <- mammal_trees[[1]]$tip.label
keep_tips <- sort(sample(all_tips, 60))

ref_trees <- lapply(mammal_trees[1:100], ape::drop.tip,
                     tip = setdiff(all_tips, keep_tips))
class(ref_trees) <- "multiPhylo"
tipLabels <- ref_trees[[1]]$tip.label
nTip <- length(tipLabels)
```

The reference pool has `r length(ref_trees)` trees on `r nTip` tips.  The
majority-rule consensus is only
`r round(length(as.Splits(Consensus(ref_trees, p = 0.5))) / (nTip - 3) * 100, 1)`%
resolved, indicating substantial disagreement among bootstrap replicates ---
a realistic scenario for CID consensus.

## Candidate generation

We generate 500 random single-TBR neighbours of the first reference tree
using `RootedTBRSwap()`.  This simulates the candidate set that the search
engine compares during TBR hill-climbing.

```{r}
#| label: tbr-neighbours
set.seed(3761)
base <- Preorder(ref_trees[[1]])
parent <- base$edge[, 1]
child <- base$edge[, 2]
nEdge <- length(parent)

tbr_neighbours <- vector("list", 500L)
for (i in seq_len(500L)) {
  result <- RootedTBRSwap(parent, child, nEdge)
  tbr_neighbours[[i]] <- structure(list(
    edge = cbind(result[[1]], result[[2]]),
    tip.label = tipLabels,
    Nnode = nTip - 1L
  ), class = "phylo")
}
class(tbr_neighbours) <- "multiPhylo"
```

RF distances from the base tree to neighbours range from
`r min(RobinsonFoulds(base, tbr_neighbours[1:50]))`
to `r max(RobinsonFoulds(base, tbr_neighbours[1:50]))`,
confirming these are genuine single-move neighbours (a TBR move can
change multiple splits simultaneously).

## MRP matrix

Each non-trivial split from each of the 100 reference trees becomes a
binary MRP character (tip in split $\to$ state 1, else $\to$ state 0).

```{r}
#| label: mrp-matrix
all_splits <- lapply(ref_trees, as.Splits, tipLabels)
mrp_logical <- do.call(rbind, lapply(all_splits, as.logical))
mrp_df <- as.data.frame(ifelse(mrp_logical, 1L, 0L))
colnames(mrp_df) <- tipLabels
mrp_phydat <- phyDat(mrp_df, type = "USER", levels = c(0, 1))
```

The MRP matrix has `r nrow(mrp_logical)` characters
(`r attr(mrp_phydat, "nr")` unique patterns) across `r ncol(mrp_logical)`
tips.

## Scoring

### CID scores

For each TBR neighbour, we compute mean CID to all 100 reference trees
using `ClusteringInfoDistance()`.

```{r}
#| label: cid-scores
cid_scores <- vapply(tbr_neighbours, function(cand) {
  mean(ClusteringInfoDistance(cand, ref_trees))
}, double(1))
```

### Parsimony scores

```{r}
#| label: parsimony-scores
ew_scores <- vapply(tbr_neighbours, function(cand) {
  TreeLength(cand, mrp_phydat)
}, double(1))

k_values <- c(1, 2, 3, 5, 10, 20)
iw_scores <- matrix(NA_real_, nrow = 500L, ncol = length(k_values))
colnames(iw_scores) <- paste0("k", k_values)
for (ki in seq_along(k_values)) {
  iw_scores[, ki] <- vapply(tbr_neighbours, function(cand) {
    TreeLength(cand, mrp_phydat, concavity = k_values[ki])
  }, double(1))
}

profile_scores <- vapply(tbr_neighbours, function(cand) {
  TreeLength(cand, mrp_phydat, concavity = "profile")
}, double(1))
```

## Results

```{r}
#| label: tbl-correlations
#| tbl-cap: "Spearman and Pearson correlations between parsimony score and mean CID across 500 TBR neighbours."
methods <- c("EW", paste0("IW k=", k_values), "Profile")
all_scores <- cbind(ew_scores, iw_scores, profile_scores)
colnames(all_scores) <- methods

cors <- data.frame(
  Method = methods,
  Spearman = round(vapply(methods, function(m) {
    cor(all_scores[, m], cid_scores, method = "spearman")
  }, double(1)), 4),
  Pearson = round(vapply(methods, function(m) {
    cor(all_scores[, m], cid_scores, method = "pearson")
  }, double(1)), 4),
  row.names = NULL
)
knitr::kable(cors)
```

```{r}
#| label: fig-scatter
#| fig-cap: "Standardized parsimony score vs mean CID for three representative methods."
#| fig-width: 7
#| fig-height: 5
plot_df <- data.frame(
  CID = rep(cid_scores, 3),
  Score = c(scale(ew_scores)[, 1],
            scale(iw_scores[, which(k_values == 3)])[, 1],
            scale(profile_scores)[, 1]),
  Method = rep(c("EW", "IW k=3", "Profile"), each = length(cid_scores))
)

ggplot(plot_df, aes(CID, Score, colour = Method)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  labs(x = "Mean CID to reference trees",
       y = "Parsimony score (standardized)")
```

```{r}
#| label: fig-residuals
#| fig-cap: "Residual parsimony score (after removing CID trend) by method. Tighter distributions indicate better proxies."
#| fig-width: 7
#| fig-height: 4
resid_df <- do.call(rbind, lapply(methods, function(m) {
  fit <- lm(all_scores[, m] ~ cid_scores)
  data.frame(Method = m, Residual = scale(residuals(fit))[, 1])
}))

ggplot(resid_df, aes(Method, Residual)) +
  geom_boxplot(outlier.size = 0.5) +
  coord_flip() +
  labs(y = "Standardized residual (parsimony | CID)")
```

## Discussion

_To be completed after results are available._

## Session info

```{r}
#| label: session-info
sessionInfo()
```
