---
title: "Tree search with custom optimality criteria"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa-old-doi-prefix.csl
vignette: >
  %\VignetteIndexEntry{Tree search with custom optimality criteria}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Using custom optimality criteria

> **For standard parsimony searches** (equal weights, implied weights, or
> profile parsimony), use `MaximizeParsimony()` — it is much faster and
> easier. See the [tree search vignette](tree-search.html).
> The functions below (`TreeSearch()`, `Ratchet()`) are for **custom
> optimality criteria only**.

"TreeSearch" can be used to search for trees that are optimal under 
user-specified criteria [e.g. @Hopkins2021].

All that is needed is to provide a function that will return a score for a
given tree (and dataset).

First we'll load the necessary R libraries:

```{r init, message = FALSE}
library("TreeTools", quietly = TRUE, warn.conflict = FALSE)
library("TreeSearch")

# Plot trees nicely
PlotTree <- function(tree, ...) {
  oPar <- par(mar = rep(0, 4), cex = 0.9)
  plot(tree)
  par(oPar)
}
```

## Maximizing tree balance

We'll start with a very simple case: aiming to minimise the 
total cophenetic index (TCI) [@Mir2013] of a tree. 
The TCI is a measure of tree balance; it does not depend on an input dataset.
All we need to do is to write a `TreeScorer` function.  
This function will be sent the `parent` and `child` nodes for each edge in a
tree, and a (here, empty) `dataset` parameter.  The function should return
a score to be minimized.  
Here, we can use a copy of our starting tree as a template, to be populated
with the rearranged `parent` and `child` vectors:

```{r tci-setup}
tree <- PectinateTree(8)
PlotTree(tree)

TCIScore <- function(parent, child, dataset) {
  tree$edge <- cbind(parent, child)
  TotalCopheneticIndex(tree)
}

TCIScore(tree$edge[, 1], tree$edge[, 2], NA)
```

Now we can use our scorer for tree search.
`dataset` is passed to `TreeScorer()` unchanged, so we can send it whatever our
scorer needs -- here, `EmptyPhyDat(tree)`, a blank dataset, as `TCIScore()`
doesn't use it at all.

```{r tci-search}
result <- TreeSearch(tree, dataset = EmptyPhyDat(tree),
                     TreeScorer = TCIScore,
                     maxIter = 50L, maxHits = 10L,
                     verbosity = 1L)

PlotTree(result)
```

## Maximizing tree distance

Let's make things slightly more complex, and try to find the tree that is 
most different from a starting tree.  Notice that `TreeSearch` aims to 
_minimize_ the output of `TreeScorer()`, so we negate the tree _distance_ 
(which we aim to maximize) before returning it.

```{r cid}
startTree <- BalancedTree(8)

DistanceScore <- function(parent, child, dataset) {
  tmpTree <- startTree
  tmpTree$edge <- cbind(parent, child)
  distance <- TreeDist::ClusteringInfoDistance(startTree, tmpTree)
  # Return:
  -distance
}

result <- TreeSearch(RandomTree(8, root = TRUE), dataset = EmptyPhyDat(tree),
                     TreeScorer = DistanceScore,
                     maxIter = 50L, maxHits = 10L,
                     verbosity = 1L)

par(mfrow = c(1, 2))
PlotTree(startTree)
PlotTree(result)

```



## More sophisticated searches

Real searches may involve more sophisticated interaction with an underlying
dataset. To demonstrate this workflow, we consider how we would implement
implied weighting as a custom search.
(This is not intended to match the efficiency of the native
`MaximizeParsimony(dataset, concavity = k)` implementation)

Implied weights [@Goloboff1993] penalize additional homoplasy according to a
concavity constant _k_: the fit contributed by a character is
`e / (e + k)`, where `e` is the number of steps in excess of the character's
theoretical minimum.  This minimum can be computed once before starting the
search, then cached; `e` depends on the tree, so is calculated once per
evaluation.


```{r iw-score, message = FALSE}
data("inapplicable.datasets")
dataset <- congreveLamsdellMatrices[[42]]

# Calculate the minimum length once; IWScore() below closes over this value,
# so it is looked up (cheaply) rather than recomputed on every call.
minLength <- MinimumLength(dataset, compress = TRUE)

IWScore <- function (parent, child, dataset, concavity = 10) {
  tree <- structure(list(edge = cbind(parent, child),
                        tip.label = names(dataset),
                        Nnode = length(dataset) - 1L), class = "phylo")
  homoplasy <- CharacterLength(tree, dataset, compress = TRUE) - minLength
  fit <- homoplasy / (homoplasy + concavity)
  # Return:
  sum(fit * attr(dataset, "weight"))
}

iwTree <- TreeSearch(NJTree(dataset), dataset, TreeScorer = IWScore,
                     maxIter = 50L, maxHits = 10L, verbosity = 1)
```

The parsimony ratchet [@Nixon1999] can help to escape local optima.
This needs a `Bootstrapper()` that resamples `dataset`'s characters: as
`dataset` here is a plain `phyDat` object, its weights live in the `"weight"`
attribute (contrast this with the native implementation's `BootstrapTree()`,
which resamples the `weight` element of a `ParsimonyData` object -- the same
idea, applied to a different data structure):

```{r iw-bootstrap}
IWBootstrap <- function (edgeList, dataset, EdgeSwapper = NNISwap,
                         maxIter, maxHits, verbosity = 1L, ...) {
  startWeights <- attr(dataset, "weight")
  resampling <- tabulate(sample(rep.int(seq_along(startWeights), startWeights),
                                replace = TRUE), length(startWeights))
  # R copy-on-modify: the caller's `dataset` is unchanged.
  attr(dataset, "weight") <- as.integer(resampling)

  res <- EdgeListSearch(edgeList[1:2], dataset, TreeScorer = IWScore,
                        EdgeSwapper = EdgeSwapper,
                        maxIter = maxIter, maxHits = maxHits,
                        verbosity = verbosity - 1L, ...)
  # Return:
  res[1:2]
}
```

```{r iw-ratchet, message = FALSE}
ratchetTree <- Ratchet(iwTree, dataset, TreeScorer = IWScore,
                       Bootstrapper = IWBootstrap,
                       ratchIter = 2, ratchHits = 2,
                       searchIter = 20, searchHits = 10,
                       verbosity = 2)
```

Use much larger values of `ratchIter`, `ratchHits`, `searchIter` and
`searchHits` for a thorough search.

Hopefully these examples give a template from which you are able to construct
your own optimality criteria.  The maintainer is happy to answer questions via
e-mail, or you can file queries by opening a
[GitHub issue](https://github.com/ms609/TreeDist/issues/new/).

## What next?

You might want to:

- [Load data](https://ms609.github.io/TreeTools/articles/load-data.html) from a Nexus file or spreadsheet

- Conduct parsimony search using Brazeau, Guillerme & Smith's [approach to inapplicable data](tree-search.html), or using [Profile parsimony](profile.html).


See also: 

- [Guide to installation](getting-started.html)

- [Documentation home](https://ms609.github.io/TreeSearch/)

- [Mapping the space of optimal trees](tree-space.html)


## References