---
title: "Tree search with custom optimality criteria"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa-old-doi-prefix.csl
vignette: >
  %\VignetteIndexEntry{Tree search with custom optimality criteria}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Using custom optimality criteria

> **For standard parsimony searches** (equal weights, implied weights, or
> profile parsimony), use `MaximizeParsimony()` — it is much faster and
> easier. See the [tree search vignette](tree-search.html).
> The functions below (`TreeSearch()`, `Ratchet()`) are for **custom
> optimality criteria and also accept `concavity` for implied weights
> or profile parsimony searches with manual control over search parameters.

"TreeSearch" can be used to search for trees that are optimal under 
user-specified criteria [e.g. @Hopkins2021].

All that is needed is to provide a function that will return a score for a
given tree (and dataset).

First we'll load the necessary R libraries:

```{r init, message = FALSE}
library("TreeTools", quietly = TRUE, warn.conflict = FALSE)
library("TreeSearch")

# Plot trees nicely
PlotTree <- function(tree, ...) {
  oPar <- par(mar = rep(0, 4), cex = 0.9)
  plot(tree)
  par(oPar)
}
```

## Maximizing tree balance

We'll start with a very simple case: aiming to minimise the 
total cophenetic index (TCI) [@Mir2013] of a tree. 
The TCI is a measure of tree balance; it does not depend on an input dataset.
All we need to do is to write a `TreeScorer` function.  
This function will be sent the `parent` and `child` nodes for each edge in a
tree, and a (here, empty) `dataset` parameter.  The function should return
a score to be minimized.  
Here, we can use a copy of our starting tree as a template, to be populated
with the rearranged `parent` and `child` vectors:

```{r tci-setup}
tree <- PectinateTree(8)
PlotTree(tree)

TCIScore <- function(parent, child, dataset) {
  tree$edge <- cbind(parent, child)
  TotalCopheneticIndex(tree)
}

TCIScore(tree$edge[, 1], tree$edge[, 2], NA)
```

Now we can use our scorer for tree search.
We need to initialize some parameters to `TreeSearch()` with null values:
`dataset = EmptyPhyDat(tree)` sends a blank dataset (as our tree scorer doesn't
require any data); we set `InitializeData = DoNothing` and
`CleanUpData = DoNothing` because we don't need to do anything to `dataset`
before it is sent to `TreeScorer()`.

```{r tci-search}
result <- TreeSearch(tree, dataset = EmptyPhyDat(tree),
                     InitializeData = DoNothing, CleanUpData = DoNothing,
                     TreeScorer = TCIScore,
                     maxIter = 50L, maxHits = 10L, 
                     verbosity = 1L)

PlotTree(result)
```

## Maximizing tree distance

Let's make things slightly more complex, and try to find the tree that is 
most different from a starting tree.  Notice that `TreeSearch` aims to 
_minimize_ the output of `TreeScorer()`, so we negate the tree _distance_ 
(which we aim to maximize) before returning it.

```{r cid}
startTree <- BalancedTree(8)

DistanceScore <- function(parent, child, dataset) {
  tmpTree <- startTree
  tmpTree$edge <- cbind(parent, child)
  distance <- TreeDist::ClusteringInfoDistance(startTree, tmpTree)
  # Return:
  -distance
}

result <- TreeSearch(RandomTree(8, root = TRUE), dataset = EmptyPhyDat(tree),
                     InitializeData = DoNothing, CleanUpData = DoNothing,
                     TreeScorer = DistanceScore,
                     maxIter = 50L, maxHits = 10L, 
                     verbosity = 1L)

par(mfrow = c(1, 2))
PlotTree(startTree)
PlotTree(result)

```



## Searching using implied weights

> **Note:** For real implied weights analyses, use
> `MaximizeParsimony(dataset, concavity = k)`, which is faster and
> uses the full driven search pipeline (see the
> [tree search vignette](tree-search.html#implied-weighting)).
> The example below shows how `TreeSearch()` and `Ratchet()` support
> implied weights directly via the `concavity` parameter.

Implied weights [@Goloboff1993] penalize additional homoplasy according to
the concavity constant _k_.  `TreeSearch()` and `Ratchet()` accept
`concavity` directly---no custom scorer or initialization code is needed:

```{r iw-search, message = FALSE}
data("inapplicable.datasets")
dataset <- congreveLamsdellMatrices[[42]]

iwTree <- TreeSearch(NJTree(dataset), dataset, concavity = 10,
                     maxIter = 50L, maxHits = 10L, verbosity = 1)
```

The parsimony ratchet [@Nixon1999] can help to escape local optima.
Bootstrap resampling is handled automatically:

```{r iw-ratchet, message = FALSE}
ratchetTree <- Ratchet(iwTree, dataset, concavity = 10,
                       ratchIter = 2, ratchHits = 2,
                       searchIter = 20, searchHits = 10,
                       verbosity = 2)
```

Use much larger values of `ratchIter`, `ratchHits`,
`searchIter` and `searchHits` for a thorough search.
Profile parsimony [@Faith2001] is also supported: set
`concavity = "profile"`.

Hopefully these examples give a template from which you are able to construct
your own optimality criteria.  The maintainer is happy to answer questions via
e-mail, or you can file queries by opening a
[GitHub issue](https://github.com/ms609/TreeSearch/issues/new/).

## What next?

You might want to:

- [Load data](https://ms609.github.io/TreeTools/articles/load-data.html) from a Nexus file or spreadsheet

- Conduct parsimony search using Brazeau, Guillerme & Smith's [approach to inapplicable data](tree-search.html), or using [Profile parsimony](profile.html).


See also: 

- [Guide to installation](getting-started.html)

- [Documentation home](https://ms609.github.io/TreeSearch/)

- [Mapping the space of optimal trees](tree-space.html)


## References
