---
title: "Inapplicable characters"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa-old-doi-prefix.csl
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inapplicable characters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Morphological datasets frequently contain characters that are "inapplicable"
for certain taxa.
For example, a character describing tail colour is inapplicable for taxa that
lack a tail; one describing wing venation is inapplicable for wingless taxa.
These characters are coded with the token `"-"` in the data matrix.

TreeSearch provides three approaches to handling inapplicable characters,
selected via the `inapplicable` argument to `MaximizeParsimony()`.

## Approaches

### Brazeau et al. (default)

The default method [@Brazeau2019] uses a three-pass Fitch-like algorithm
that infers applicability regions from the `"-"` token.
No additional information is needed:

```{r brazeau, eval = FALSE}
MaximizeParsimony(dataset)
```

This approach does not require explicit specification of which characters
depend on which, because it infers inapplicable regions directly from the
`"-"` coding.

### Hopkins & St. John (HSJ)

The HSJ method [@Hopkins2021] scores inapplicable characters using a
dissimilarity-metric approach.
Each hierarchy block—consisting of a controlling primary character
(presence/absence) and its dependent secondary characters—is scored
with a dynamic programming algorithm that accounts for secondary
character variation within the "present" region.

The `hsj_alpha` parameter (0–1) controls how much weight is given to
secondary character variation:

- `hsj_alpha = 0`: Only the primary absent/present pattern matters.
- `hsj_alpha = 1` (default): Each branch where the primary is present at both
  endpoints incurs a penalty proportional to the fraction of secondary
  characters that differ.

```{r hsj, eval = FALSE}
hierarchy <- CharacterHierarchy("1" = 2:5, "6" = 7:8)
MaximizeParsimony(dataset, hierarchy = hierarchy,
                  inapplicable = "hsj", hsj_alpha = 1.0)
```

### X-transformation (step-matrix recoding)

The x-transformation approach [@Goloboff2021] recodes each hierarchy block
as a single step-matrix character with asymmetric costs.
The "absent" state is separated from all possible combinations of
secondary character states, with a high cost for gaining the structure
(absent → present) and a low cost for losing it (present → absent):

- **Gain cost**: _n_ + 1 (where _n_ is the number of secondary characters)
- **Loss cost**: 1
- **Present → present**: Hamming distance (number of secondary characters that differ)

This asymmetry captures the idea that independently evolving a complex
structure (and its associated secondary characters) is less parsimonious
than losing it.

```{r xform, eval = FALSE}
hierarchy <- CharacterHierarchy("1" = 2:5)
MaximizeParsimony(dataset, hierarchy = hierarchy,
                  inapplicable = "xform")
```

## Specifying the character hierarchy

Both the HSJ and x-transformation methods require a `CharacterHierarchy`
object that specifies which characters control which.

### Manual specification

Use `CharacterHierarchy()` with named arguments.
Each name is the index of a controlling (primary) character,
and each value is a vector of the character indices it controls:

```{r hierarchy-manual}
library(TreeSearch)

# Character 1 controls characters 2-5
h <- CharacterHierarchy("1" = 2:5)

# Multiple controlling primaries
h <- CharacterHierarchy("1" = 2:5, "6" = 7:8)
```

### Auto-detection from character names

If your dataset uses the TNT naming convention (`sup_<tag>` for controlling
characters, `sub_<tag>_<suffix>` for their dependents), you can auto-detect
the hierarchy:

```{r hierarchy-auto}
names <- c("sup_tail", "sub_tail_colour", "sub_tail_shape",
           "sup_wing", "sub_wing_venation", "other_char")
hierarchy_from_names(names)
```

## Worked example

```{r example}
# Create a small dataset with inapplicable characters
mat <- matrix(c(
  # pri  sec1  sec2  indep1  indep2
  "0",  "-",  "-",  "0",    "0",
  "0",  "-",  "-",  "0",    "1",
  "1",  "0",  "0",  "1",    "0",
  "1",  "0",  "1",  "1",    "0",
  "1",  "1",  "0",  "1",    "1",
  "1",  "1",  "1",  "0",    "1"
), nrow = 6, byrow = TRUE,
dimnames = list(paste0("t", 1:6), NULL))
dataset <- TreeTools::MatrixToPhyDat(mat)
hierarchy <- CharacterHierarchy("1" = 2:3)

# Default: Brazeau et al. algorithm (no hierarchy needed)
brazeau_trees <- MaximizeParsimony(dataset, maxReplicates = 3,
                                   targetHits = 2, verbosity = 0)

# HSJ: dissimilarity-metric scoring
hsj_trees <- MaximizeParsimony(dataset, hierarchy = hierarchy,
                               inapplicable = "hsj", hsj_alpha = 1.0,
                               maxReplicates = 3, targetHits = 2,
                               verbosity = 0)

# X-transformation: step-matrix recoding
xform_trees <- MaximizeParsimony(dataset, hierarchy = hierarchy,
                                 inapplicable = "xform",
                                 maxReplicates = 3, targetHits = 2,
                                 verbosity = 0)
```

## Choosing a method

| Criterion | Brazeau | HSJ | X-transformation |
|-----------|---------|-----|------------------|
| Hierarchy required | No | Yes | Yes |
| Secondary char. variation | Not modelled | Penalised (via \U{03B1}) | Penalised (via Hamming) |
| Gain/loss asymmetry | No | No | Yes (n+1 : 1) |
| Tuning parameter | None | `hsj_alpha` | None |
| Speed | Fastest | Moderate | Moderate |

**Brazeau** is appropriate when you do not have a clear hierarchy
specification or want a fast default analysis.

**HSJ** is appropriate when you have a hierarchy and want to
penalize secondary character variation, with control over how
strongly secondaries contribute (via `hsj_alpha`).

**X-transformation** is appropriate when you believe gains of
complex structures should be penalized more heavily than losses,
following the maximum-homology philosophy of @Goloboff2021.

## References
