---
title: "Tree Search with Profile Parsimony"
author: "Martin R. Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
csl: ../inst/apa-old-doi-prefix.csl
vignette: >
  %\VignetteIndexEntry{Tree Search with Profile Parsimony}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Profile Parsimony [@Faith2001] finds the tree that is most faithful 
to the information contained within a given dataset. 
It is the 'exact solution' that implied weights parsimony approximates.

Profile Parsimony is currently implemented in 'TreeSearch' only 
for binary characters with no ambiguous tokens.

## Getting started
<!--Duplicated from inapplicable.Rmd-->
[A companion vignette](getting-started.html) gives details on installing 
the package and getting up and running.

<!--
# message=FALSE will suppress message when loading TreeSearch
# Temporary fix until R.oo is updated -- remove thereafter
# https://github.com/r-lib/rlang/issues/669
-->

Once installed, load the inapplicable package into R using
```{r Load-library, message=FALSE}
library('TreeSearch')
```

In order to reproduce the random elements of this document, set a random seed:

```{R RNG-version}
# Set a random seed so that random functions in this document are reproducible
suppressWarnings(RNGversion("3.5.0")) # Until we can require R3.6.0
set.seed(888)
```


## Scoring a tree, and conducting a tree search

Here's an example of using the package to conduct tree search with profile 
parsimony.
You can [load your own dataset](https://ms609.github.io/TreeTools/articles/load-data.html), but for this example,
we'll use a simulated dataset that comes bundled with the `TreeSearch` package.

```{r Load Longrich data}
data(congreveLamsdellMatrices)
myMatrix <- congreveLamsdellMatrices[[10]]
```

To start analysis, we need to load a starting tree.  We can do this at random:
```{r Random tree, eval=FALSE}
tree <- TreeTools::RandomTree(myMatrix, root = TRUE)
```

Or using a neighbour joining method, to start at a reasonably good tree:
```{r NJ Tree}
tree <- TreeTools::NJTree(myMatrix)
par(mar = rep(0.25, 4), cex = 0.75) # make plot easier to read
plot(tree)
```

Let's calculate this random tree's parsimony score,
then search for better trees using the parsimony ratchet [@Nixon1999]:

```{R Starting score}
TreeLength(tree, myMatrix, 'profile')

betterTrees <- MaximizeParsimony(myMatrix, tree, concavity = 'profile',
                                  ratchIter = 3, tbrIter = 3, maxHits = 8)
```

We've used very low values of `ratchIter`, `tbrIter` and `maxHits` for a rapid
run, so this is not necessarily a thorough enough search to find a globally
optimal tree.
Nevertheless, let's see the resultant tree, and its score:

```{r Ratchet-search-results}
TreeLength(betterTrees[[1]], myMatrix, 'profile')
par(mar = rep(0.25, 4), cex = 0.75) # make plot easier to read
plot(ape::consensus(betterTrees))
```

The default parameters may not be enough to find the optimal tree; type 
`?MaximizeParsimony` to view all search parameters --
or keep repeating the search until tree score stops improving.

## View the results

In parsimony search, it is good practice to consider trees that are slightly suboptimal [@Smith2019].

Here, we'll take a consensus that includes all trees that are suboptimal by up
to 3 bits.
To sample this region of tree space well, the trick is to use large values of 
`ratchHits` and `ratchIter`, and small values of `searchHits` and
`searchiter`, so that many runs don't quite hit the optimal tree.
In a serious study, you would want to sample many more than the 3 Ratchet hits (`ratchHits`) we'll settle for here, probably using many more Ratchet iterations.

```{R Suboptimal sampling}
suboptimals <- MaximizeParsimony(myMatrix, betterTrees, tolerance = 3,
                                 ratchIter = 2, tbrIter = 3,
                                 maxHits = 25,
                                 concavity = 'profile')
```

The consensus of these slightly suboptimal trees provides a less resolved, but
typically more reliable, summary of the signal with the phylogenetic dataset
[@Smith2019]:

```{r Plot suboptimal consensus}
par(mar = rep(0.25, 4), cex = 0.75)
table(TreeLength(suboptimals, myMatrix, 'profile'))
plot(ape::consensus(suboptimals))
```


## References

