---
title: "TreeSearch vs TNT: Benchmark Comparison"
author: "TreeSearch Development"
date: today
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

## Overview
  
This report compares the phylogenetic search performance of
**TreeSearch** (C++ driven search engine) against **TNT 1.6**
(Goloboff & Catalano 2016) across a suite of morphological datasets.

Both programs implement "new technology" search strategies (Wagner tree +
TBR + sectorial search + ratchet + drift + tree fusing). The comparison
focuses on:

- **Score quality**: Does each engine find equally good trees?
- **Speed**: Wall-clock time to convergence.
- **Scaling**: How does performance change with dataset size?

### Scoring caveat

TreeSearch uses the Brazeau et al. (2019) three-pass algorithm for
inapplicable characters, while TNT treats inapplicable tokens as
a regular character state. For datasets with inapplicable data, the two
programs optimise *different objective functions*, so absolute scores are
not directly comparable. What is comparable is each program's ability to
find its own optimum efficiently.

```{r}
#| label: setup
library(tidyverse)
library(knitr)
```

## Load results

```{r}
#| label: load-data

# Find the most recent results CSV
csv_files <- list.files(".", pattern = "^tnt_compare_.*\\.csv$",
                        full.names = TRUE)
if (length(csv_files) == 0) {
  stop("No results CSV found. Run bench_tnt_compare.R first.")
}
results_file <- sort(csv_files, decreasing = TRUE)[1]
results <- read.csv(results_file, stringsAsFactors = FALSE)
cat("Loaded", nrow(results), "rows from", basename(results_file), "\n")
```

## Score comparison

For each dataset, we compare the best score found across all seeds.

```{r}
#| label: score-summary

score_summary <- results |>
  group_by(dataset, n_taxa, n_chars, weighting) |>
  summarise(
    tnt_best = min(tnt_score, na.rm = TRUE),
    ts_best = min(ts_score, na.rm = TRUE),
    tnt_median = median(tnt_score, na.rm = TRUE),
    ts_median = median(ts_score, na.rm = TRUE),
    n_runs = n(),
    .groups = "drop"
  ) |>
  arrange(n_taxa)

kable(score_summary, digits = 2,
      caption = "Best and median scores per dataset and weighting scheme")
```

## Timing comparison

```{r}
#| label: timing-summary

timing_summary <- results |>
  group_by(dataset, n_taxa, weighting, timeout_s) |>
  summarise(
    tnt_median_s = median(tnt_wall_s, na.rm = TRUE),
    ts_median_s = median(ts_wall_s, na.rm = TRUE),
    speedup = median(tnt_wall_s, na.rm = TRUE) /
      median(ts_wall_s, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(n_taxa)

kable(timing_summary, digits = 3,
      caption = "Median wall-clock time (seconds) and TNT/TreeSearch ratio")
```

### Timing by dataset size

```{r}
#| label: fig-timing
#| fig-cap: "Wall-clock time vs number of taxa"
#| fig-width: 8
#| fig-height: 5

results_long <- results |>
  select(dataset, n_taxa, weighting, timeout_s, seed,
         tnt_wall_s, ts_wall_s) |>
  pivot_longer(cols = c(tnt_wall_s, ts_wall_s),
               names_to = "engine",
               values_to = "wall_s") |>
  mutate(engine = ifelse(engine == "tnt_wall_s", "TNT", "TreeSearch"))

ggplot(results_long, aes(n_taxa, wall_s, colour = engine)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~weighting) +
  labs(x = "Number of taxa", y = "Wall-clock time (s)",
       colour = "Engine") +
  theme_minimal()
```

## Score quality

How often does each engine find the best score (across both engines)?

```{r}
#| label: score-quality

quality <- results |>
  group_by(dataset, weighting) |>
  mutate(
    overall_best = min(c(tnt_score, ts_score), na.rm = TRUE),
    tnt_optimal = tnt_score <= overall_best * 1.001,
    ts_optimal = ts_score <= overall_best * 1.001
  ) |>
  summarise(
    n_runs = n(),
    tnt_pct_optimal = round(100 * mean(tnt_optimal, na.rm = TRUE), 1),
    ts_pct_optimal = round(100 * mean(ts_optimal, na.rm = TRUE), 1),
    .groups = "drop"
  )

kable(quality,
      caption = "Percentage of runs finding the best-known score (within 0.1%)")
```

## Score vs taxa scatter

```{r}
#| label: fig-score-scatter
#| fig-cap: "Best scores found by each engine per dataset"
#| fig-width: 8
#| fig-height: 5

score_long <- score_summary |>
  select(dataset, n_taxa, weighting, tnt_best, ts_best) |>
  pivot_longer(cols = c(tnt_best, ts_best),
               names_to = "engine", values_to = "best_score") |>
  mutate(engine = ifelse(engine == "tnt_best", "TNT", "TreeSearch"))

ggplot(score_long, aes(n_taxa, best_score, colour = engine)) +
  geom_point(size = 2.5) +
  geom_line(aes(group = interaction(dataset, weighting)), alpha = 0.3,
            colour = "grey50") +
  facet_wrap(~weighting, scales = "free_y") +
  labs(x = "Number of taxa", y = "Best score",
       colour = "Engine") +
  theme_minimal()
```

## Rearrangements (TNT)

TNT reports the total number of rearrangements examined, which gives a
sense of search intensity.

```{r}
#| label: tnt-rearrangements

if ("tnt_rearrangements" %in% names(results)) {
  rearr <- results |>
    group_by(dataset, n_taxa, weighting, timeout_s) |>
    summarise(
      median_rearrangements = median(tnt_rearrangements, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(n_taxa)
  
  kable(rearr, digits = 0,
        caption = "Median TNT rearrangements per dataset")
}
```

## Raw results

```{r}
#| label: raw-results

kable(results |> arrange(dataset, weighting, timeout_s, seed),
      digits = 3,
      caption = "Full results table")
```

## Session info

```{r}
#| label: session-info
sessionInfo()
```
